---
title: 'TCGA Differential Expression Analysis: Tumor vs Normal'
author: "Thomas Norburn"
date: "2025-10-16"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# 0) Configuration — set your paths here

> Define the base directory once. All downstream steps reference this
> variable.

```{r config}
base_dir <- "C:/Users/tno/Documents/Coding"   # forward slashes work on Windows in R
rdata_file <- file.path(base_dir, "tcga_luad.RData")
```

# 1) Download required packages and data

Relevant packages (if not already downloaded)

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c(
  'TCGAbiolinks', 'GO.db', 'DESeq2', 'edgeR', 'limma', 'pheatmap',
  'ComplexHeatmap', 'clusterProfiler', 'org.Hs.eg.db', 'AnnotationDbi',
  'biomaRt', 'EnhancedVolcano', 'survival', 'survminer',
  'enrichplot', 'ReactomePA'
))
# Note: clusterProfiler appears once (was duplicated in the original)
```

Load packages

```{r libraries, message=FALSE, warning=FALSE}
library(DESeq2)
library(data.table)
library(edgeR)
library(limma)
library(pheatmap)
library(ComplexHeatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(biomaRt)
library(EnhancedVolcano)
library(survival)
library(survminer)
library(enrichplot)
library(ReactomePA)
library(TCGAbiolinks)
```

## Download TCGA-LUAD data

```{r download, eval=FALSE}
query <- GDCquery(
  project       = 'TCGA-LUAD',
  data.category = 'Transcriptome Profiling',
  data.type     = 'Gene Expression Quantification',
  workflow.type = 'STAR - Counts'
)
GDCdownload(query, directory = base_dir)
data <- GDCprepare(query, directory = base_dir)
```

# 2) Save and reload prepared data

> Save the prepared SummarizedExperiment object for faster reloading in
> future sessions.

```{r save-data, eval=FALSE}
# FIX: rdata_file points to a file, not a directory
save(data, file = rdata_file)
```

> Once saved, load directly to skip re-downloading:

```{r load-data}
load(rdata_file)
```

# 3) Extract counts and clinical data

> We start by extracting the expression counts and clinical metadata
> from the TCGA dataset. This separates the numeric expression data from
> patient/sample-level information, which is necessary for downstream
> analyses.

```{r extract}
counts   <- assay(data)
clinical <- colData(data)
```

## Standardise column and row names

> Why:
>
> -   Ensure unique column names for proper alignment with clinical
>     data.
> -   Remove version numbers from Ensembl IDs (e.g. ENSG00000141510.18 →
>     ENSG00000141510).
> -   Restore TCGA-style hyphenated sample IDs after `make.names()`
>     converts hyphens to dots.

```{r clean-names}
# Step 1: make column names syntactically valid and unique
colnames(counts) <- make.names(colnames(counts), unique = TRUE)

# Step 2: strip Ensembl version numbers from row names
rownames(counts) <- sub("\\..*", "", rownames(counts))

# Step 3: restore TCGA hyphen-delimited sample IDs
colnames(counts) <- gsub("\\.", "-", colnames(counts))
```

## Check sample type distribution

```{r sample-types}
colnames(clinical)
table(clinical$sample_type)
```

## Create Tumor / Normal group variable

```{r group-factor}
clinical$group <- ifelse(
  clinical$sample_type == "Solid Tissue Normal", "Normal", "Tumor"
)
# Make it a factor so DESeq2 uses Normal as the reference level
clinical$group <- factor(clinical$group, levels = c("Normal", "Tumor"))
```

## Align counts and clinical sample IDs

> Ensures counts matrix and clinical metadata are in the same order,
> which is required by DESeq2.

```{r align-samples}
common_samples <- intersect(colnames(counts), rownames(clinical))
counts         <- counts[, common_samples]
clinical       <- clinical[common_samples, ]
stopifnot(all(colnames(counts) == rownames(clinical)))  # must be TRUE

table(clinical$group)
```

# 4) Quality Control and filtering

> Remove genes with low counts across samples (CPM \< 1). Normalise
> using TMM (edgeR) to account for library size differences. Prepare
> log-transformed CPM for exploratory analysis.

```{r qc-filter}
dge  <- DGEList(counts = counts)
keep <- filterByExpr(dge, group = clinical$group)
dge  <- dge[keep, , keep.lib.sizes = FALSE]
dge  <- calcNormFactors(dge)   # TMM normalisation
logCPM <- cpm(dge, log = TRUE)
```

## PCA for quality check

> PCA visualises sample clustering based on global gene expression. Used
> to detect potential batch effects or outliers before DE analysis.

```{r pca-edger}
pr <- prcomp(t(logCPM))
plot(
  pr$x[, 1:2],
  col  = ifelse(clinical$group == "Tumor", "red", "blue"),
  pch  = 16,
  xlab = "PC1", ylab = "PC2",
  main = "PCA of logCPM (edgeR TMM-normalised)"
)
legend("topright", legend = c("Tumor", "Normal"), col = c("red", "blue"), pch = 16)
```

# 5) Differential expression analysis (DESeq2)

> DESeq2 is a robust method for RNA-seq DE analysis. Goal: identify
> genes significantly up- or down-regulated between tumour and normal
> tissues.

## Build DESeq2 dataset

```{r deseq2-build}
counts_filtered <- counts[keep, ]
dds <- DESeqDataSetFromMatrix(
  countData = counts_filtered,
  colData   = clinical,
  design    = ~ group
)
```

## Run DESeq2

```{r deseq2-run}
dds        <- DESeq(dds)
res        <- results(dds, contrast = c("group", "Tumor", "Normal"))
resOrdered <- res[order(res$padj), ]
```

## Save DESeq2 results to disk

```{r deseq2-save}
dir.create("results", showWarnings = FALSE)
write.csv(as.data.frame(resOrdered), file = "results/deseq2_results2.csv")
```

## PCA on DESeq2 VST-normalised counts

> Confirms sample grouping after variance-stabilising transformation.
> `vsd` is stored and reused in later steps to avoid redundant
> computation.

```{r pca-vst}
vsd <- vst(dds)
plotPCA(vsd, intgroup = "group")
```

## Annotate results with HGNC gene symbols

```{r annotate}
# Strip any remaining Ensembl version numbers
rownames(resOrdered) <- sub("\\..*", "", rownames(resOrdered))

# Map Ensembl IDs to HGNC symbols
resOrdered$symbol <- mapIds(
  org.Hs.eg.db,
  keys     = rownames(resOrdered),
  column   = "SYMBOL",
  keytype  = "ENSEMBL",
  multiVals = "first"
)

# Fall back to Ensembl ID where no symbol is available
resOrdered$symbol[is.na(resOrdered$symbol)] <-
  rownames(resOrdered)[is.na(resOrdered$symbol)]
```

## Summary of significant results

```{r deseq2-summary}
# Separate chunk so this renders clearly in the report
summary(res$padj < 0.05)
head(resOrdered)
```

## Label top 10 up- and down-regulated *significant* genes for volcano plot

> FIX: labels now drawn only from genes that pass the padj \< 0.05
> threshold, preventing statistically non-significant genes from being
> highlighted.

```{r label-genes}
resOrdered$label <- ""

# Restrict to statistically significant genes before ranking by fold change
sig_for_labels <- resOrdered[
  !is.na(resOrdered$padj) & resOrdered$padj < 0.05, 
]

top_up   <- head(rownames(sig_for_labels[order(sig_for_labels$log2FoldChange, decreasing = TRUE), ]), 10)
top_down <- head(rownames(sig_for_labels[order(sig_for_labels$log2FoldChange, decreasing = FALSE), ]), 10)

label_idx <- rownames(resOrdered) %in% c(top_up, top_down)
resOrdered$label[label_idx] <- resOrdered$symbol[label_idx]
```

# 6) Volcano plot

> Visualise the most significant and biologically meaningful DEGs.
> Labels highlight the top 10 up- and down-regulated significant genes.

```{r volcano}
EnhancedVolcano(
  resOrdered,
  lab              = resOrdered$label,
  x                = "log2FoldChange",
  y                = "padj",
  pCutoff          = 0.05,
  FCcutoff         = 1,
  title            = "Differential Expression: Tumor vs Normal",
  subtitle         = "TCGA-LUAD",
  caption          = "DESeq2 analysis",
  pointSize        = 2.5,
  labSize          = 3.0,
  col              = c("blue", "yellow", "black", "red"),
  colAlpha         = 1,
  drawConnectors   = TRUE,
  widthConnectors  = 0.5,
  gridlines.major  = FALSE,
  gridlines.minor  = FALSE
)
```

# 7) Heatmap of top 50 DEGs

> Heatmap shows expression patterns across all samples for the top 50
> DEGs. Scaled rows allow relative expression comparison across samples.

```{r heatmap-top50}
# Use a distinct variable name to avoid conflict with the data frame used in Section 8
topGenes_vec  <- rownames(head(resOrdered, 50))
norm_counts   <- assay(vsd)   # reuse vsd computed in Section 5
mat           <- norm_counts[topGenes_vec, ]

pheatmap(
  mat,
  scale          = "row",
  annotation_col = data.frame(Group = clinical$group, row.names = colnames(mat))
)
```

# 8) Correlation among top 20 DEGs

> Explores co-regulation among the most significant DEGs. Spearman
> correlation is used as it is robust to outliers and non-normality.

```{r correlation-heatmap}
# FIX: uses distinct variable name (topGenes_df) rather than reusing topGenes_vec
topGenes_df  <- head(resOrdered[order(resOrdered$padj), ], 20)

# FIX: reuses vsd from Section 5 instead of calling vst(dds) a second time
exprMatrix   <- assay(vsd)[rownames(topGenes_df), ]
rownames(exprMatrix) <- topGenes_df$symbol

corMatrix <- cor(t(exprMatrix), method = "spearman")

pheatmap(
  corMatrix,
  clustering_distance_rows = "correlation",
  clustering_distance_cols = "correlation",
  main     = "Spearman correlation among top 20 DE genes",
  fontsize = 10
)
```

# 9) Pathway enrichment and annotation

> Filter significant DEGs (adjusted p-value \< 0.05) and run GO, KEGG,
> and Reactome enrichment to interpret DE results in a biological
> context.

## Prepare significant gene list

```{r sig-genes}
sig_genes <- resOrdered[
  which(!is.na(resOrdered$padj) & resOrdered$padj < 0.05), 
]
```

## Convert to Entrez IDs

> Most pathway databases (KEGG, Reactome, GO) use Entrez Gene IDs.
> `mapIds()` converts from Ensembl. `multiVals = "first"` handles
> many-to-one mappings cleanly.

```{r entrez-ids}
sig_genes$entrez <- mapIds(
  org.Hs.eg.db,
  keys      = rownames(sig_genes),
  column    = "ENTREZID",
  keytype   = "ENSEMBL",
  multiVals = "first"
)

sig_entrez <- na.omit(sig_genes$entrez)

# FIX: guard against empty gene list before running enrichment
if (length(sig_entrez) == 0) {
  stop("No significant genes mapped to Entrez IDs. Check padj threshold or gene annotation.")
}

message(length(sig_entrez), " genes mapped to Entrez IDs for enrichment analysis.")
```

## Ranked gene list for GSEA

> A ranked gene list (by log2FC) is required for Gene Set Enrichment
> Analysis. Unlike ORA, GSEA considers all genes and is less sensitive
> to arbitrary p-value cutoffs, making it better at detecting
> coordinated pathway changes.

```{r gsea-list}
geneList <- resOrdered$log2FoldChange
names(geneList) <- mapIds(
  org.Hs.eg.db,
  keys      = rownames(resOrdered),
  column    = "ENTREZID",
  keytype   = "ENSEMBL",
  multiVals = "first"
)
geneList <- sort(na.omit(geneList), decreasing = TRUE)
```

## GO Biological Process enrichment

> `enrichGO()` identifies overrepresented GO terms (Biological Process)
> among significant genes. BH correction accounts for multiple testing.
> `readable = TRUE` converts Entrez IDs back to gene symbols.

```{r go-enrichment}
ego_bp <- enrichGO(
  gene          = sig_entrez,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

head(ego_bp, 10)
```

```{r go-plots}
dotplot(ego_bp, showCategory = 15) + ggtitle("GO BP Enrichment (Tumor vs Normal)")
barplot(ego_bp, showCategory = 15)
```

## KEGG pathway enrichment

> `enrichKEGG()` identifies signalling and metabolic pathways from KEGG.
> `organism = "hsa"` specifies human. `setReadable()` converts IDs to
> gene symbols for interpretation.

```{r kegg-enrichment}
ekegg <- enrichKEGG(
  gene         = sig_entrez,
  organism     = "hsa",
  keyType      = "kegg",
  pvalueCutoff = 0.05
)
ekegg <- setReadable(ekegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

dotplot(ekegg, showCategory = 15) + ggtitle("KEGG Pathway Enrichment (Tumor vs Normal)")
```

## Reactome pathway enrichment

> `enrichPathway()` uses the manually curated Reactome database, which
> is more granular than KEGG and covers signal transduction, immune
> response, and gene regulation pathways relevant to cancer.

```{r reactome-enrichment}
react <- enrichPathway(
  gene         = sig_entrez,
  organism     = "human",
  pvalueCutoff = 0.05,
  readable     = TRUE
)

head(react, 10)
dotplot(react, showCategory = 15) + ggtitle("Reactome Pathway Enrichment (Tumor vs Normal)")
```

# 10) Save enrichment results

> FIX: all three write.csv() calls are now on single lines with valid
> paths. Single forward slashes work on all platforms in R.

```{r save-enrichment}
dir.create("results/enrichment", showWarnings = FALSE, recursive = TRUE)

write.csv(as.data.frame(ego_bp), file = "results/enrichment/GO_BP.csv")
write.csv(as.data.frame(ekegg),  file = "results/enrichment/KEGG.csv")
write.csv(as.data.frame(react),  file = "results/enrichment/Reactome.csv")

message("Enrichment results saved to results/enrichment/")
```

# 11) Session information

> Always include sessionInfo() in published or shared bioinformatics
> reports. Records exact R and package versions for full
> reproducibility.

```{r session-info}
sessionInfo()
```
